\chapter{\sffamily Empirical dynamical emulators}

{\bfseries\sffamily Concept.} To leverage the probabilistic formalism that we developed in the previous chapter to be able to empirically emulate the real-world data for a wide variety stochastic phenomena within a generalised framework. This chapter will then explore how the mathematics fits as code within the stochadex and follows this up with some simple examples. For the mathematically-inclined, this chapter will take a detailed look at how our formalism can be adapted to focus on dynamical process emulation. For the programmers, we will be creating some new modules in the stochadex public Git repository: \href{https://github.com/umbralcalc/stochadex}{https://github.com/umbralcalc/stochadex}.

\section{\sffamily Probabilistic formalism}

The key distinction between the methods that we will develop in this chapter and the ones in the next chapter (on statistical inference tools) is in their utility when faced with the problem of attempting to model real-world data. In the proceeding chapter, we shall describe some powerful techniques that can be used most effectively when the researcher is aware of the family of models that generated the data. In the present chapter, we will go into the details of how a more 'empirical' approach can be derived for dynamical process modelling in a probabilistic framework which locally adapts the model to the data through time. This technique enables the researcher to model complex trends in the data dynamically very well; at the cost of making the abstract interpretation of the model less immediately comprehensible.

To begin with we want to...

\href{https://en.wikipedia.org/wiki/Empirical_dynamic_modeling#:~:text=Empirical%20dynamic%20modeling%20(EDM)%20is,Robert%20May%20and%20George%20Sugihara.}{reference this}

\href{https://en.wikipedia.org/wiki/Gaussian_process#:~:text=In%20probability%20theory%20and%20statistics,of%20them%20is%20normally%20distributed.}{and this}

\href{https://en.wikipedia.org/wiki/Local_regression}{and also this}